A Stopped Negative Binomial Distribution

### 2016-04-07 stat-and-prob-letters/snb-paper.tex
Updated based on Dan's comments.

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. If it is a success, the process 
advances one diagonally in the
positive horizontal and vertical direction.
Otherwise, it advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

@@

Next, define
\begin{equation} \label{stop_t}
S'(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
and notice that $S(k, p, s) = S'(k, 1-p, s)$ by writing
\begin{equation*}
{k-1 \choose k-s} = {k-1 \choose s-1}.
\end{equation*}

To show that (\ref{eqn:N}) and (\ref{stop_t}) sum to one
over their support, let
\begin{align} \label{eqn:sum_proof}
R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{align}
where we substitute $i=k-s$ in the first summation and $j=k-t$ in the second.

@@

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010} and recall this function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Abramowitz1964}.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is a valid probability mass
function.
\end{proof}

\begin{corollary} \label{conditional_distribution}
Assume a trial that stops on the $s^{th}$ success or $t^{th}$ failure as 
before. 
Furthermore, assume we observer $s'$ successes and $t'$ failures
where $s' < s$ and $t' < t$.
That is, we see the number of success and failures at a time before
reaching an endpoint. Then the number of subsequent enrollements needed 
to reach an endpoint is SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
\begin{proof}
The result follows directly by realizing that, conditioned on $s'$
success and $t'$ failures, there are $s-s'$ successes needed to reach the
success endpoint and $t-t'$ failures needed to reach the failure endpoint.
\end{proof}

\section{Shape and Basic Properties}

@@

\section{Connection Between the SNB and the Binomial Tail Probability}

The probability of reaching the success endpoint in an 
SNB($p$, $s$, $t$) random variable is 
equal to the probability of at least $s$ successes in a Binomial distributed 
random variable with size $s+t-1$ and success probability $p$. 
Likewise, the probability of reaching the failure endpoint is equal
to at most $s-1$ successes in a Binomial distribution.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n\ |\ \text{\#success} = s].
\end{equation}
That is, the probability that the number of successes is at least $s$
in the Binomial model is the same that the trial stops with $s$ 
successes in the SNB model.
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
  & 1 - \mathcal{I}_{1-p}.
\end{align*}
The corresponding SNB probability is
\begin{align*}
\mathbb{P} [Y \leq n\ |\ \text{\#success}=s] &= \sum_{k=s}^{s+t-1} S(k, p, s)\\
 &= {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{align*}
Let $i=k-s$. Use ${i+s-1 \choose s-1} = {i+s-1 \choose i}$
so the last summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n\ |\ \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

Fig.~\ref{fig:snb_bin_compare} shows the case
where $s=2$, $t=12$, and $p=0.2$.  The probability masses represented in 
black are equal in a) and b) as are the masses in grey.
The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

@@

The moment generating function is calculated in a manner similar to 
that of two negative binomial distributions. However, in the negative
binomial derivation the
term involving the regularized incomplete beta function integrates to
one since it is integrated over it's entire support. This is not the
case in the SNB and it appears in both terms.

\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ when $x \leq \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose k-s} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose k-s} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one, we have
$qe^x \leq 1$. This also shows $x \leq -\log(q)$.

A similar expression can be derived using the same calculation with
the constraint that $x \leq -\log(p)$. The result
follows from the aforementioned property of the regularized incomplete
beta function.
\end{proof}

@@

Let us consider a Bayesian analysis where there is a Beta prior distribution
on $p$.
\begin{prop} \label{prop:bayesian}
The posterior PMF of the Stopped Negative Binomial distribution with a Beta($\alpha$, $\beta$) prior is:
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $s,t \leq k \leq s+t-1$. When this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}

@@

The SNB can be used to extend clinical trial methodology
in at least two areas. First, it can be used to construct nested criteria
for early stopping and efficacy. These trials consist of two-stages.
If the failure endpoint is reached in the first stage then the 
trial stops due to futility. Otherwise, the trial enters the second 
stage, which consists of responders from the first stage along with
new enrollees in the second. Thus the design provides
both early testing for futility along with the ability
to make statistically sound decisions sooner than traditional trials.

A second avenue for methodological innovation comes directly from
Corollary \ref{conditional_distribution} and Proposition \ref{prop:bayesian}.
While a trial is being conducted the response probabilities can be fitted
and updated as new results are received. The number of individual successes
and failure are used to learn the shape characteristics of the corresponding
Beta distribution and resulting posterior
distribution. An estimate of the probability of reaching the success (and
failure) endpoints can be provided at each step of the trial.

There are also several avenues for extending the SNB itself. First is the 
development of a stopped negative multinomial distribution allowing
for more than two endpoints. This extension would have clinical trial
applications including dose escalation where the outcomes are response
to a therapy, failure to respond, and an adverse outcome endpoint where
an enrollee is removed from the trial with neither response or non-response
due to side effects.

A second, more theoretical extension is to the asymptotic setting. Under
proper reparameterization, the process from which the SNB is derived 
can be shown to converge to a Brownian motion. In this case, we conjecture
that the SNB can be defined in terms of the L\'{e}vy-Bachelier 
\citep{Bachelier1900} formula
thereby relating the SNB to the classic result in stochastic calculus.

>>>>

To prove (\ref{eqn:pmf}), consider the
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ I_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. In Fig.~\ref{fig:kane_viz} 
we consider a graphical illustration of the plot $X_k$ against
$k$. If the outcome of the $k$th patient is a success then the process 
advances one diagonally in the
positive horizontal and vertical direction. If the $k$th patient fails
then the sample path advances in the positive horizontal direction only. The
process continues until either $X_k = s$ or $X_k = k-t$.

@@

%Next, define
%\begin{equation} \label{stop_t}
%S'(k, p, t) = {k-1 \choose k-t} p^{k-t} (1-p)^t
%\end{equation}
%and notice that $S(k, p, s) = S'(k, 1-p, s)$ by writing
%\begin{equation*}
%{k-1 \choose k-s} = {k-1 \choose s-1}.
%\end{equation*}

To show that Equation \ref{eqn:pmf} sums to one, define
\begin{align} \label{eqn:sum_proof}
R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{align}
where we substitute $i=k-s$ in the first summation and $j=k-t$ in the second.

@@

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010} and recall this function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Abramowitz1964}.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof (\ref{eqn:pmf}) is is the distribution of the
stopping time and a valid probability mass
function.
\end{proof}

Next, we consider an interim analysis of a clinical trial after $s'$ 
successes and $t'$ failures have been observed for $s' < s$ and $t' < t$.
\begin{corollary} \label{conditional_distribution}
The number of subsequent enrollements needed 
to reach an endpoint is SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
Having observed $s'$
successes and $t'$ failures, there are $s-s'$ successes needed to reach the
success endpoint and $t-t'$ failures needed to reach the failure endpoint.

\section{Connections and Approximations by Other Distributions}

@@

%\section{Connection Between the SNB and the Binomial Tail Probability}

There is a close connection between the tail areas of the SNB and the 
binomial distributions.
The probability of reaching the success endpoint in an 
SNB($p$, $s$, $t$) random variable is 
equal to the probability of at least $s$ successes in a Binomial distributed 
random variable with size $s+t-1$ and success probability $p$. 
Likewise, the probability of reaching the failure endpoint is equal
to at most $s-1$ successes in a Binomial distribution. That is, the
probability that the number of successes is at least $s$ in the Binomial
model is the same that the trial succeeds (reaches $s$) in the SNB model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let 
$B$ be distributed Binomial with size $n=s+t-1$ and success probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [Y \leq n\ |\ \text{trial succeeds}].
\end{equation}
\end{prop}
\begin{proof}
The Binomial tail probability is
\begin{align*}
\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
  &= 1 - \mathcal{I}_{1-p}.
\end{align*}
The corresponding SNB probability is
\begin{equation*}
\mathbb{P} [Y \leq n\ |\ \text{trial succeeds}=s] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation*}
Let $i=k-s$. Using the fact that
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i}
\end{equation*}
the last summation can be rewritten as
\begin{align}
\mathbb{P} [Y \leq n\ |\ \text{\#success} = s] &= \sum_{i=0}^{t-1} 
  {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align}
completing the proof.
\end{proof}

To illustrate this result, let us return to our initial example, shown
in Fig.~\ref{fig:snb_bin_compare}, 
where $s=2$, $t=12$, and $p=0.2$.  The probability masses represented in 
black are equal in a) and b) as are the masses in grey.
The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

@@

The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. 
\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ when $x \leq \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose k-s} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
The first summation in Equation \ref{eqn:first_sum} satisfies
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose k-s} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose k-s} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function's subscript parameter has support on zero 
to one, we have $qe^x \leq 1$. This also shows $x \leq -\log(q)$.

A similar expression can be derived for the second summation in 
Equation \ref{eqn:first_sum} and results in
the constraint $x \leq -\log(p)$.
\end{proof}

@@

Let us consider a Bayesian analysis with a Beta prior distribution on $p$.
\begin{prop} \label{prop:bayesian}
The posterior PMF of the Stopped Negative Binomial distribution with a 
Beta($\alpha$, $\beta$) prior is
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $\min(s,t) \leq k \leq s+t-1$. When 
this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp.
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}

@@

The SNB can be used to extend clinical trial methodology
in at least two areas. First, it can be used to construct nested criteria
for early stopping and efficacy. Current approaches are often inspired by
the Simon 2-stage Phase II clinical trial \cite{Simon1989} where a
a small number of patients are enrolled for an initial stage. For patients,
success or failure is assessed. If at least a pre-specified number of
successes are observed then the trial enters a second stage with more
patients. If a sufficient number of successes are not observed then the
trial fails.

The SNB is being used to generalize the Simon design by providing separate
criteria for futility and efficacy.
The Simon design tests for a single response probability and time to event
parameter for both the first
and second stage. The SNB is being used to generalize this design by 
allowing these parameters to vary for the two stages, and even allow
information about patients in the first stage to be used in the second
stage. Early results show the design provides
both early testing for futility along with the ability
to make statistically sound decisions sooner than traditional trials.

A second avenue for methodological innovation comes directly from
Corollary \ref{conditional_distribution} and Proposition \ref{prop:bayesian}.
Our knowledge of the response probability $p$ comes from published reports
yielding a prior distribution.
While a trial is being conducted the response probabilities are 
updated as new results are received. The number of individual successes
and failure are used to learn the shape characteristics of the corresponding
Beta distribution and resulting posterior
distribution. An estimate of the probability of reaching the success (and
failure) endpoints can be provided at each step of the trial.

%There are also several avenues for extending the SNB itself. First is the 
%development of a stopped negative multinomial distribution allowing
%for more than two endpoints. This extension would have clinical trial
%applications including dose escalation where the outcomes are response
%to a therapy, failure to respond, and an adverse outcome endpoint where
%an enrollee is removed from the trial with neither response or non-response
%due to side effects.

%A second, more theoretical extension is to the asymptotic setting. Under
%proper reparameterization, the process from which the SNB is derived 
%can be shown to converge to a Brownian motion. In this case, we conjecture
%that the SNB can be defined in terms of the L\'{e}vy-Bachelier 
%\citep{Bachelier1900} formula
%thereby relating the SNB to the classic result in stochastic calculus.
